---
title: "Flexible dose-response models"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```



```{r include=FALSE}
### Libraries and analysis objects

# The master script runs the current pooled PAROXETINE analysis and creates
# all objects displayed below.
source("00_master.R")

if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("Package not installed: patchwork")
}
library(patchwork)

make_plot_grid <- function(plot_list, ncol = 1) {
  plot_list <- purrr::compact(plot_list)
  if (length(plot_list) == 0) return(NULL)
  patchwork::wrap_plots(plot_list, ncol = ncol)
}

print_kable_list <- function(tab_list, caption, digits = 3) {
  for (nm in names(tab_list)) {
    print(
      knitr::kable(
        as.data.frame(tab_list[[nm]]),
        digits = digits,
        caption = paste0(
          "<small><em>",
          caption,
          ": ",
          nm,
          "</em></small>"
        )
      )
    )
    cat("\n\n")
  }
}

stack_named_tables <- function(tab_list) {
  purrr::imap_dfr(
    tab_list,
    function(tab, nm) {
      as.data.frame(tab) |>
        dplyr::mutate(
          analysis_name = nm,
          .before = 1
        )
    }
  )
}

PLOT_STRATEGY_DOSES <- c(20, 30, 40, 50)

analysis_summary <- analysis_zero_as_dose |>
  dplyr::summarise(
    n_active_patients = dplyr::n_distinct(pid),
    n_active_rows = dplyr::n(),
    n_msm_patients = dplyr::n_distinct(pid[use_msm]),
    n_msm_rows = sum(use_msm, na.rm = TRUE),
    dose_levels = paste(
      levels(droplevels(dose_f)),
      collapse = ", "
    )
  )

study_summary <- analysis_zero_as_dose |>
  dplyr::group_by(studyid) |>
  dplyr::summarise(
    n_active_patients = dplyr::n_distinct(pid),
    n_active_rows = dplyr::n(),
    .groups = "drop"
  )

dose_by_window_grid <- make_plot_grid(
  dose_by_window_plots,
  ncol = 1
)

dose_transition_grid <- make_plot_grid(
  dose_transition_plots,
  ncol = 1
)

dose_grouping_grid <- make_plot_grid(
  dose_grouping_plots,
  ncol = 1
)

iptw_dose_model_grid <- make_plot_grid(
  iptw_dose_model_plots,
  ncol = 1
)

iptw_response_profile_grid <- make_plot_grid(
  iptw_response_profile_plots,
  ncol = 1
)

ipcw_denominator_grid <- make_plot_grid(
  ipcw_denominator_plots,
  ncol = 1
)

multinom_dose_model_grid <- if (
  exists("multinom_dose_model_plots")
) {
  make_plot_grid(
    multinom_dose_model_plots,
    ncol = 1
  )
} else {
  NULL
}

dose_vs_placebo_table_readme <- dose_vs_placebo_predictions_all |>
  dplyr::filter(
    strategy_dose %in% PLOT_STRATEGY_DOSES
  )

gee_and_placebo_means_table_readme <- gee_and_placebo_means_all |>
  dplyr::filter(
    is.na(strategy_dose) |
      strategy_dose %in% PLOT_STRATEGY_DOSES
  )

if (exists("multinom_dose_vs_placebo_predictions_all")) {
  multinom_dose_vs_placebo_table_readme <-
    multinom_dose_vs_placebo_predictions_all |>
    dplyr::filter(
      strategy_dose %in% PLOT_STRATEGY_DOSES
    )
}

if (exists("compare_ordinal_vs_multinom_placebo")) {
  ordinal_vs_multinom_table_readme <-
    compare_ordinal_vs_multinom_placebo |>
    dplyr::filter(
      strategy_dose %in% PLOT_STRATEGY_DOSES
    )
}
```


## Clinical outcome, safety measure, time-varying confounders, and dataset

The clinical outcome is the Hamilton Depression Rating Scale (HAMD) score, denoted by \(Y_{it}\), measured for patient \(i\) at observation \(t\). The baseline HAMD score is denoted by \(Y_{i0}\). Treatment response is expressed as improvement from baseline:

```math
\Delta Y_{it} = Y_{i0} - Y_{it}.
```

Thus, positive values of \(\Delta Y_{it}\) indicate improvement.

The current analysis pools all eligible studies containing both **PAROXETINE** and placebo. The primary analysis retains actual reported study time and expresses it continuously in weeks. Follow-up for the outcome analysis is restricted to day 42 (6 weeks), while the complete available follow-up is retained to determine whether a patient truly had no later observation (i.e., permanent loss to follow-up before the analysis horizon).

The observed PAROXETINE dose is treated as a time-varying treatment. Zero dose is retained as a genuine observed treatment state in the treatment-assignment and dose-history models. Sparse dose support is grouped before modelling; in the current data 10 mg is represented as 20 mg and 60 mg as 50 mg, while the original reported dose is preserved separately. The clinically presented sustained strategies are 20, 30, 40, and 50 mg; the 0 mg strategy is not displayed in the final figures.

Safety is summarized using a side-effect score \(S_{it}\) on a 0--10 scale, denoting the number of relevant to PAROXETINE side effects reported. In the current analysis these values are simulated.

Previous dose \(d_{it-1}\) is part of the treatment process because it defines the dose trajectory and strongly predicts the next assigned dose. The post-baseline HAMD improvement \(\Delta Y_{it}\) and side-effect severity \(S_{it}\) are treated as time-varying confounders because they may be affected by previous treatment and may also influence subsequent dose assignment, future outcome, or censoring.

```{r summary-efficacy-safety, echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(
  analysis_summary,
  col.names = c(
    "Number of active patients",
    "Number of active patient-visits",
    "Number of MSM patients",
    "Number of MSM patient-visits",
    "Observed dose levels (mg)"
  ),
  caption = "<small><em>Pooled PAROXETINE analysis summary</em></small>"
)

knitr::kable(
  study_summary,
  col.names = c(
    "Study",
    "Number of active patients",
    "Number of active patient-visits"
  ),
  caption = "<small><em>Active-arm contribution by study</em></small>"
)
```


### Empirical support and positivity diagnostics

Before interpreting the weighted marginal structural model, we assessed empirical support for the observed dose process. Positivity requires that, within the histories represented in the analysis, the dose categories being compared have adequate observed support. Very sparse dose levels, empty transition cells, or near-deterministic dose assignment can produce unstable treatment weights and unreliable extrapolation.

Broad time windows are used only for descriptive summaries and diagnostic plots. All statistical models retain continuous study time.

```{r dose-by-window-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
dose_by_window_grid
```

```{r dose-transition-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
dose_transition_grid
```

```{r dose-grouping-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
dose_grouping_grid
```

```{r analysis-modelability, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  analysis_modelability,
  caption = "<small><em>Support check for the pooled active analysis</em></small>"
)
```


## Notation

<table class="notation-table">
<thead>
<tr>
<th>Notation</th>
<th>Explanation</th>
</tr>
</thead>
<tbody>

<tr>
<td>\(i = 1,\ldots,n\)</td>
<td>Patient index.</td>
</tr>

<tr>
<td>\(t\)</td>
<td>Ordered observation/decision index for patient \(i\); baseline is the first observation.</td>
</tr>

<tr>
<td>\(\mathrm{week}_{it}\)</td>
<td>Actual continuous study week of observation \(t\) for patient \(i\).</td>
</tr>

<tr>
<td>\(g_i\)</td>
<td>Study identifier for patient \(i\), entered as a fixed effect.</td>
</tr>

<tr>
<td>\(Y_{it}\)</td>
<td>HAMD score for patient \(i\) at observation \(t\).</td>
</tr>

<tr>
<td>\(\Delta Y_{it} = Y_{i0} - Y_{it}\)</td>
<td>HAMD improvement from baseline. Positive values indicate improvement.</td>
</tr>

<tr>
<td>\(S_{it}\)</td>
<td>Side-effect score for patient \(i\) at observation \(t\).</td>
</tr>

<tr>
<td>\(d_{it}\)</td>
<td>Observed dose assigned at observation \(t\).</td>
</tr>

<tr>
<td>\(R_{it+1}\)</td>
<td>Indicator of remaining uncensored after observation \(t\). It is 0 only when no later observation exists in the full available follow-up before the 6-week horizon; it is not defined after the analysis horizon.</td>
</tr>

<tr>
<td>\(\hat p_{it}^{D}\)</td>
<td>Fitted denominator probability of the observed dose.</td>
</tr>

<tr>
<td>\(\hat q_{it}^{D}\)</td>
<td>Fitted numerator probability of the observed dose.</td>
</tr>

<tr>
<td>\(\mathrm{SW}_{it}^{D}\)</td>
<td>Visit-specific stabilized dose weight.</td>
</tr>

<tr>
<td>\(\mathrm{cSW}_{it}^{D}\)</td>
<td>Cumulative stabilized dose weight applied to the outcome at observation \(t\), based on treatment decisions before that outcome.</td>
</tr>

<tr>
<td>\(\hat p_{it}^{C}\)</td>
<td>Fitted denominator probability of remaining uncensored after observation \(t\).</td>
</tr>

<tr>
<td>\(\hat q_{it}^{C}\)</td>
<td>Fitted numerator probability of remaining uncensored after observation \(t\).</td>
</tr>

<tr>
<td>\(\mathrm{SW}_{it}^{C}\)</td>
<td>Visit-specific stabilized censoring weight.</td>
</tr>

<tr>
<td>\(\mathrm{cSW}_{it}^{C}\)</td>
<td>Cumulative stabilized censoring weight applied to the outcome at observation \(t\), based on preceding censoring intervals.</td>
</tr>

<tr>
<td>\(\mathrm{SW}_{it}^{\mathrm{total}}\)</td>
<td>Final stabilized treatment-by-censoring weight.</td>
</tr>

</tbody>
</table>


## Step 1 - Inverse probability of dose and censoring weighting (IPDCW)

### Inverse probability of dose weights (IPTW)

The first step is to construct stabilized inverse probability of dose weights to adjust for time-varying confounding affected by prior dose, following Robins, Hernán and Brumback (2000).

The **denominator model** represents the full observed dose-assignment mechanism. The **numerator model** is a reduced model used to stabilize the weights. Both models include study as a fixed effect because the analysis pools patients from several trials.

#### Dose weight denominator

The primary dose-weight denominator is fitted at the patient-observation level using ordinal logistic regression. Dose is an ordered categorical variable with levels \(0 < 20 < 30 < 40 < 50\) mg. For each cumulative threshold \(c\), the model is

```math
\begin{aligned}
\mathrm{logit}
\left\{
\Pr(D_{it} \le c \mid H_{it})
\right\}
&=
\alpha_{0c}
+ \alpha_1 f(\mathrm{week}_{it})
+ \alpha_2 g_i
+ \alpha_3 \Delta Y_{it}
+ \alpha_4 S_{it}
+ \alpha_5 d_{it-1} \\
&\quad
+ \alpha_6 d_{it-1}\Delta Y_{it}
+ \alpha_7 d_{it-1}S_{it}
+ \alpha_8 Y_{i0}.
\end{aligned}
```

The restricted cubic spline \(f(\mathrm{week}_{it})\) allows the dose-assignment process to vary non-linearly over actual continuous study time. Previous dose is represented categorically. The interactions with previous dose allow the same efficacy or side-effect information to lead to different dose decisions depending on the dose already received.

The denominator probability used in the weight is the fitted probability of the dose actually assigned:

```math
\hat p_{it}^{D}
=
\widehat{\Pr}
\left(
D_{it}=d_{it}
\mid
H_{it}
\right).
```

```{r iptw-denominator-coef-tabs, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
print_kable_list(
  iptw_denominator_coef_tabs,
  caption = "Coefficient table for the ordinal dose-weight denominator",
  digits = 3
)
```

The observed-versus-predicted diagnostic compares the empirical proportion assigned to each dose with the mean model-predicted probability, within study and broad time window.

```{r iptw-dose-model-observed-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=13, fig.height=9}
iptw_dose_model_grid
```

The response-profile diagnostic is descriptive. It compares fitted dose-assignment probabilities among observed patient-visits with different response histories while setting the side-effect score to the same value in the displayed profiles. It should not be interpreted as a randomized manipulation of response.

```{r iptw-response-profile-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}
iptw_response_profile_grid
```


#### Dose weight numerator

The numerator excludes the time-varying confounders while retaining continuous time, study, previous dose, and baseline HAMD:

```math
\begin{aligned}
\mathrm{logit}
\left\{
\Pr(D_{it} \le c \mid H_{it}^{*})
\right\}
&=
\beta_{0c}
+ \beta_1 f(\mathrm{week}_{it})
+ \beta_2 g_i
+ \beta_3 d_{it-1}
+ \beta_4 Y_{i0}.
\end{aligned}
```

The fitted probability corresponding to the observed dose is

```math
\hat q_{it}^{D}
=
\widehat{\Pr}
\left(
D_{it}=d_{it}
\mid
H_{it}^{*}
\right).
```

```{r iptw-numerator-coef-tabs, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
print_kable_list(
  iptw_numerator_coef_tabs,
  caption = "Coefficient table for the ordinal dose-weight numerator",
  digits = 3
)
```


#### Stabilized dose weights and timing

The visit-specific stabilized dose weight is

```math
\mathrm{SW}_{it}^{D}
=
\frac{
\hat q_{it}^{D}
}{
\hat p_{it}^{D}
}.
```

HAMD and side effects are observed at a visit before the dose prescribed at that visit is used for the next treatment interval. Therefore the treatment weight estimated from the dose decision at observation \(t\) is applied to **subsequent** outcomes, not to the outcome already measured at observation \(t\). The cumulative treatment weight attached to outcome \(t\) is therefore

```math
\mathrm{cSW}_{it}^{D}
=
\prod_{s < t}
\mathrm{SW}_{is}^{D}.
```

This temporal alignment differs from multiplying the current decision weight into the same-row outcome.

```{r iptw-treatment-weight-summary-clean, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  stack_named_tables(iptw_weight_summaries),
  digits = 3,
  caption = "<small><em>Summary of visit-specific and cumulative stabilized dose weights</em></small>"
)
```

```{r iptw-positivity, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  iptw_probability_diagnostics,
  digits = 3,
  caption = "<small><em>Ordinal treatment-model positivity diagnostics</em></small>"
)
```

The ordinal model assumes proportional odds: covariates shift the probability toward higher or lower dose categories similarly across cumulative dose thresholds. This assumption is examined in the Appendix using a multinomial treatment model.


### Inverse probability of censoring weights (IPCW)

Censoring is defined using all available follow-up, not merely the absence of an observation at the next nominal week. For an observation before week 6, \(R_{it+1}=0\) only when the patient has no later observation anywhere in the complete available follow-up. Thus irregular visit spacing by itself does not define censoring.

#### Censoring weight denominator

The denominator model includes study, a restricted cubic spline of continuous time, current dose, current and previous HAMD improvement, current and previous side-effect scores, baseline HAMD, age, and sex:

```math
\begin{aligned}
\mathrm{logit}
\left\{
\Pr(R_{it+1}=1 \mid H_{it}^{C})
\right\}
&=
\gamma_0
+ \gamma_1 f(\mathrm{week}_{it})
+ \gamma_2 g_i
+ \gamma_3 d_{it}
+ \gamma_4 \Delta Y_{it}
+ \gamma_5 \Delta Y_{it-1} \\
&\quad
+ \gamma_6 S_{it}
+ \gamma_7 S_{it-1}
+ \gamma_8 Y_{i0}
+ \gamma_9 \mathrm{age}_i
+ \gamma_{10} \mathrm{sex}_i.
\end{aligned}
```

```{r ipcw-denominator-coef-tabs, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
print_kable_list(
  ipcw_denominator_coef_tabs,
  caption = "Coefficient table for the censoring-weight denominator",
  digits = 3
)
```

```{r ipcw-observed-predicted-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}
ipcw_denominator_grid
```


#### Censoring weight numerator

The numerator excludes the time-varying efficacy and side-effect histories but retains study, continuous time, current dose, baseline HAMD, age, and sex:

```math
\begin{aligned}
\mathrm{logit}
\left\{
\Pr(R_{it+1}=1 \mid H_{it}^{C*})
\right\}
&=
\delta_0
+ \delta_1 f(\mathrm{week}_{it})
+ \delta_2 g_i
+ \delta_3 d_{it}
+ \delta_4 Y_{i0}
+ \delta_5 \mathrm{age}_i
+ \delta_6 \mathrm{sex}_i.
\end{aligned}
```

```{r ipcw-numerator-coef-tabs, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
print_kable_list(
  ipcw_numerator_coef_tabs,
  caption = "Coefficient table for the censoring-weight numerator",
  digits = 3
)
```


#### Stabilized censoring weights and timing

The visit-specific censoring weight is

```math
\mathrm{SW}_{it}^{C}
=
\frac{
\hat q_{it}^{C}
}{
\hat p_{it}^{C}
}.
```

As with treatment assignment, censoring after observation \(t\) can affect later outcomes but cannot affect the outcome already observed at \(t\). The cumulative censoring weight applied to outcome \(t\) is therefore

```math
\mathrm{cSW}_{it}^{C}
=
\prod_{s < t}
\mathrm{SW}_{is}^{C}.
```

```{r ipcw-censoring-weight-summary, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  stack_named_tables(ipcw_weight_summaries),
  digits = 3,
  caption = "<small><em>Summary of stabilized censoring weights</em></small>"
)
```


### Total stabilized weights and truncation

The final observation-level weight is

```math
\mathrm{SW}_{it}^{\mathrm{total}}
=
\mathrm{cSW}_{it}^{D}
\times
\mathrm{cSW}_{it}^{C}.
```

The total weight, rather than its treatment and censoring components separately, is truncated at the 1st and 99th percentiles. Weight stability is assessed using the distribution of the weights and the Kish effective sample size,

```math
\mathrm{ESS}
=
\frac{
\left(
\sum_{i,t}
\mathrm{SW}_{it}^{\mathrm{total}}
\right)^2
}{
\sum_{i,t}
\left(
\mathrm{SW}_{it}^{\mathrm{total}}
\right)^2
}.
```

```{r total-weight-summary, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  stack_named_tables(total_weight_summaries),
  digits = 3,
  caption = "<small><em>Summary of untruncated total stabilized weights</em></small>"
)
```

```{r total-truncated-weight-summary, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  stack_named_tables(truncated_weight_summaries),
  digits = 3,
  caption = "<small><em>Summary of total stabilized weights after 1st/99th percentile truncation</em></small>"
)
```


## Step 2 - Weighted repeated-measures marginal structural model MSM

After estimating the total weights, we fit a weighted repeated-measures marginal structural model using generalized estimating equations with an identity link and Gaussian working variance. An independence working correlation structure is used, with robust sandwich standard errors clustered by patient.

The outcome model includes a restricted cubic spline of continuous study week, baseline HAMD, the three most recent dose-history variables, the average earlier dose, and study as a fixed effect. The most recent dose is allowed to interact with time, and baseline HAMD is also allowed to interact with time:

```math
\begin{aligned}
\Delta Y_{it}
&=
\eta_0
+ \eta_1 f(\mathrm{week}_{it})
+ \eta_2 Y_{i0}
+ \eta_3 d_{it-1}
+ \eta_4 d_{it-2}
+ \eta_5 d_{it-3} \\
&\quad
+ \eta_6 \overline{d}_{i<t-3}
+ \eta_7 Y_{i0}f(\mathrm{week}_{it})
+ \eta_8\left[d_{it-1}\times f(\mathrm{week}_{it})\right]
+ \eta_9 g_i
+ \varepsilon_{it}.
\end{aligned}
```

The recent dose variables \(d_{it-1}\), \(d_{it-2}\), and \(d_{it-3}\) are represented categorically using the analysis dose levels. Patients were untreated before study entry, so structural pre-baseline dose history is represented by 0 mg. The average earlier dose \(\overline d_{i<t-3}\) is kept continuous. Age and sex are not included in the final MSM; they are used in the censoring models.

The study fixed effect allows the mean outcome level to differ across studies while estimating a pooled PAROXETINE dose-response relation.

```{r msm-coef-tabs, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
print_kable_list(
  msm_coef_tabs,
  caption = "Weighted marginal structural model for HAMD improvement",
  digits = 3
)
```


## Step 3 - Standardized dose-strategy predictions and placebo comparison

After fitting the weighted MSM, we predict HAMD improvement at weeks 1--6 under sustained PAROXETINE strategies. All patients were untreated before study entry and the observed trials initiated PAROXETINE at 20 mg. Therefore the intervention histories used for prediction start with 20 mg at baseline. For the 20 mg strategy, treatment remains at 20 mg. For the 30, 40, and 50 mg strategies, treatment starts at 20 mg at baseline and is then set to the target dose at subsequent treatment-decision occasions.

The final figures present the clinically relevant 20, 30, 40, and 50 mg strategies. Zero dose remains part of the observed treatment process and therefore remains in the IPTW and dose-history models, but the 0 mg strategy is not displayed as a target intervention.

### What is the standardized population?

Predictions are not calculated for one artificial patient with a mean baseline HAMD score. Instead, each strategy is evaluated for every patient in the active MSM population using that patient's observed baseline HAMD and study membership. The patient-specific model predictions are then averaged.

For strategy \(a\) at week \(t\), the standardized mean is

```math
\widehat{\mu}_{a}(t)
=
\frac{1}{N}
\sum_{i=1}^{N}
\widehat{Y}_{i}^{\,a}(t).
```

Thus, **standardized** means that every treatment strategy is evaluated over the same empirical target population. In the current data this is the set of active patients contributing to the MSM. The comparison between dose strategies is therefore not driven by one strategy being evaluated in a different mix of studies or in patients with a different baseline HAMD distribution.

The counterfactual dose histories use each patient's empirical treatment-decision/visit structure before the target week, while dose itself is replaced by the specified strategy. Predictions at each target week are averaged across the full active MSM target population.

### Why the graph does not use the raw observed placebo mean

A raw observed placebo mean at an exact integer week would use only placebo patients who happened to have an observation exactly at that time. With continuously recorded visit times, these exact-week cells can be sparse, particularly in smaller studies. In addition, a simple pooled placebo mean can have a different study composition and baseline-HAMD distribution from the active MSM target population.

For this reason, all available placebo observations through week 6 are first used to fit a separate placebo GEE:

```math
\Delta Y_{it}^{P}
=
\theta_0
+ \theta_1 f(\mathrm{week}_{it})
+ \theta_2 Y_{i0}
+ \theta_3 Y_{i0}f(\mathrm{week}_{it})
+ \theta_4 g_i
+ \varepsilon_{it}.
```

The placebo model is then predicted at weeks 1--6 using the **same study membership and baseline HAMD values as the active MSM target population**, and these predictions are averaged. The black curve is therefore a **standardized placebo prediction**, estimated from the observed placebo data but transported to the same target population used for the active strategies. This makes the active-versus-placebo comparison more comparable and avoids relying on sparse exact-week placebo cells.

Pointwise 95% confidence intervals for active strategies are calculated from the robust GEE covariance matrix. The active-versus-placebo standard error combines the active and placebo prediction standard errors. These intervals do not propagate uncertainty from estimation of the treatment and censoring weights; a full bootstrap would be required for that additional uncertainty.

```{r strategy-prediction-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
strategy_prediction_plot
```

```{r dose-vs-placebo-table, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  dose_vs_placebo_table_readme,
  digits = 3,
  caption = "<small><em>Dose-strategy predictions versus standardized placebo by week</em></small>"
)
```

```{r gee-and-placebo-means-table, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  gee_and_placebo_means_table_readme,
  digits = 3,
  caption = "<small><em>Active MSM predictions and standardized placebo predictions by week</em></small>"
)
```


## References

1. J M Robins, M A Hernán, B Brumback. Marginal structural models and causal inference in epidemiology. Epidemiology. 2000 Sep;11(5):550-60. doi: 10.1097/00001648-200009000-00011.
2. Ilya Lipkovich, David H Adams, Craig Mallinckrodt, Doug Faries, David Baron, John P Houston. Evaluating dose response from flexible dose clinical trials. BMC Psychiatry. 2008 Jan 7;8:3. doi: 10.1186/1471-244X-8-3.


## Appendix

### Alternative multinomial dose model for stabilized dose weights

The primary treatment-weight model uses ordinal logistic regression and therefore assumes proportional odds. As a sensitivity analysis, we refit the treatment numerator and denominator using multinomial logistic regression. The multinomial model treats the dose categories as nominal and allows separate covariate associations for each non-reference dose.

Using 20 mg as the reference dose, the denominator model for each non-reference category \(d \in \{0,30,40,50\}\) is

```math
\begin{aligned}
\log
\left[
\frac{
\Pr(D_{it}=d \mid H_{it})
}{
\Pr(D_{it}=20 \mid H_{it})
}
\right]
&=
\alpha_{0d}
+ \alpha_{1d}f(\mathrm{week}_{it})
+ \alpha_{2d}g_i
+ \alpha_{3d}\Delta Y_{it}
+ \alpha_{4d}S_{it}
+ \alpha_{5d}d_{it-1} \\
&\quad
+ \alpha_{6d}d_{it-1}\Delta Y_{it}
+ \alpha_{7d}d_{it-1}S_{it}
+ \alpha_{8d}Y_{i0}.
\end{aligned}
```

The numerator uses the same reduced history as the ordinal numerator:

```math
\begin{aligned}
\log
\left[
\frac{
\Pr(D_{it}=d \mid H_{it}^{*})
}{
\Pr(D_{it}=20 \mid H_{it}^{*})
}
\right]
&=
\beta_{0d}
+ \beta_{1d}f(\mathrm{week}_{it})
+ \beta_{2d}g_i
+ \beta_{3d}d_{it-1}
+ \beta_{4d}Y_{i0}.
\end{aligned}
```

For each observation, the fitted probability corresponding to the actually received dose is extracted from each model and used to construct stabilized treatment weights. The same lagged treatment-weight timing, the same IPCW, the same total-weight truncation, the same MSM, the same standardized target population, and the same standardized placebo comparator are then used. Thus the sensitivity analysis changes only the functional form of the treatment-assignment model.

```{r appendix-multinomial-iptw, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
if (exists("multinom_denominator_coef_tabs")) {
  print_kable_list(
    multinom_denominator_coef_tabs,
    caption = "Multinomial dose-weight denominator model",
    digits = 3
  )
}

if (exists("multinom_numerator_coef_tabs")) {
  print_kable_list(
    multinom_numerator_coef_tabs,
    caption = "Multinomial dose-weight numerator model",
    digits = 3
  )
}

if (exists("multinom_weight_summaries")) {
  print(
    knitr::kable(
      stack_named_tables(multinom_weight_summaries),
      digits = 3,
      caption = "<small><em>Summary of multinomial stabilized treatment weights</em></small>"
    )
  )
}

if (exists("multinom_total_truncated_weight_summaries")) {
  print(
    knitr::kable(
      stack_named_tables(multinom_total_truncated_weight_summaries),
      digits = 3,
      caption = "<small><em>Summary of truncated total weights using multinomial IPTW</em></small>"
    )
  )
}
```


```{r appendix-multinomial-msm, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
if (exists("multinom_msm_coef_tabs")) {
  print_kable_list(
    multinom_msm_coef_tabs,
    caption = "Weighted MSM using multinomial IPTW",
    digits = 3
  )
}
```


### Predicted trajectories using multinomial IPTW

```{r appendix-multinomial-dose-model-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=13, fig.height=9}
multinom_dose_model_grid
```

```{r appendix-multinomial-prediction-grid, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
if (exists("multinom_strategy_prediction_plot")) {
  multinom_strategy_prediction_plot
}
```

### Comparison of ordinal and multinomial IPTW results

```{r appendix-ordinal-vs-multinomial-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
if (exists("ordinal_vs_multinom_strategy_plot")) {
  ordinal_vs_multinom_strategy_plot
}
```

```{r appendix-ordinal-vs-multinomial-comparison, echo=FALSE, message=FALSE, warning=FALSE}
if (exists("ordinal_vs_multinom_table_readme")) {
  print(
    knitr::kable(
      ordinal_vs_multinom_table_readme,
      digits = 3,
      caption = "<small><em>Comparison of dose-versus-placebo predictions using ordinal versus multinomial treatment weights</em></small>"
    )
  )
}
```
